\vspace{-1.75cm}PhD Proposal

### 2022-05-17 abstract/abstract.tex
continuo abstract

\title{\vspace{-1.75cm}Ph.D. Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}
\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{0.75ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

@@

in a large collection of individual genomes. This new reference is called a
\emph{pangenome}.  

@@

thousand individuals, in the next few years, also thanks to latest developments
in sequencing technologies, (\textbf{Next Generation Sequencing

@@

has practical and efficient linear-time solutions, if there is a proper indexing

@@

  transform (\textit{BWT})} on which the \textit{FM-index} is based. 

@@

This aligner based on the \textit{seed-and-extend paradigm}, where pattern

@@

In recent years, there has been an important development
in the field of bioinformatics. Now the 
researchers are focusing on the study of \textbf{pangenomes}, which
term was introduced
% by Tettelin
in 2005 for dealing with the comparison of genes in
bacterial species. Briefly, the \textit{pangenome} is a compact 

@@

addition to the specificity of single genome, allowing even more the use of

@@

approach that will become increasingly important even in near future.\\ 

@@

The main focus of this proposal is improving the current state of the art
of pattern matching algorithms and data structures to represent and query

@@

such data. These research field will have an impact on the ability of

@@

genomic information from pangenome graphs. 

@@

% In this context, my Ph.D. is going to be focused on the development of new

@@

\textbf{Bitvectors} are one of the most important \textit{succint data
  structures}.  
A \textit{bitvector} is an array on $n$ bits that allows two particular

@@

one present in the \textit{bitvector}. Formally, given a bitvector $B$, such

@@

for the analysis of biological sequences. An example of the use is 
tracking the runs in the run-length encoded implementations of \textit{BWT} and
\textit{PBWT}, where we put one at each head of every run, allowing fast
operations for indexing and mapping. 

@@

The \textbf{Burrows-Wheeler Transform (\textit{BWT})} was introduced
in 1994 to compress texts but it has been used widely
in bioinformatics, above all thanks to the already cited
\textit{FM-index}.

@@

type of sequences and a simple but effective strategy to compress texts is based

@@

to introduce some definitons. First of all, the main purpose of these algorithms
is to compute \textbf{Maximal Exact Matches (\textit{MEMs})} of a pattern in a
text. Given a text $T$ and a pattern $P$, a substring of the pattern $P[i :
i+l-1]$, of length $l$, is a \textit{MEM} of $P$ in $T$ if $P[i:i+l-1]$ is a
substring of $T$ but neither $P[i-1:i+l-1]$ nor $P[i:i+l]$ are, that is the
match cannot be extended in any direction. The authors state that computing
\textit{MEMs} is equivalent to computing \textbf{matching statistics
  (\textit{MS})}. Formally, the \textit{MS} of a pattern $P$ in respect to a
text $T$ is an array $M$ of pairs position/length ($pos$/$len$), $|M|=|P|$, such
that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and $P[i:i+M[i].len]$
does not occur in $T$. To compute \textit{MS}, \textit{random-access} over the
text is required  but it is also required that the text is stored in a
compressed way. For this purpose, the authors get a compressed 
representation of the text, via a \textbf{straight-line program

@@

consecutive runs of the same character This solution requires two sweep over the

@@

\textit{MS}, by a single sweep. Formally, given two positions $i$ and $j$ in a

@@

extension between two positions in the text. Summarizing, after computing the
\textit{MS array}, as stated by Bannai et al., it is possible to obtain

@@

in $SA$, the previous and next index stored in the suffix array. \\
Thanks to these works, it is
possible to perform \textit{pattern matching} 
efficiently even on long sequences of nucleotides, such as those studied in
a pangenomic context. In fact, for example, according to Rossi et al.,

@@

biology, have led to the development and of algorithms and data
structures for analyzing specific sequence data such as \textit{haplotypes} and
\textit{genotypes} in the \textbf{genotyping variants problem}. Briefly, we

@@

the genotype is their combined information. So, since 2005, publication of the

@@

example to study the outcomes of diseases. In this particular historical

@@

columns, not extendable in any direction, such that there is a substet of
identical rows).

@@

research based on it, both in terms of variant design, such as the
\textbf{multiallelic PBWT} or the \textbf{dynamic PBWT}, where the static PBWT

@@

structure via linked lists, and in terms of its use to study haplotype panels.
A first example could be found in the paper of Alanko et al., published in
2021, where the authors used the \textit{PBWT} to look for \textit{maximal
  perfect haplotype block}, within a haplotypes panel. These types of
algorithms are essential for the identification of genomic regions that show
signatures of natural selection.
Another interesting paper is the one by Williams and Mumey, published in 2020,
who also studied \textit{maximal perfect haplotype blocks}, but with the

@@

software is making \textit{genotype imputation} to predict unobserved

@@

\textit{GWAS} on tumors and other disorders.\\
To concludes, as introduced,
haplotypes panel can be thought of as a particular case of  
\textit{pangenome}, with fixed-length genomes, so we can understand the
importance of such compressed and efficient data structure in
\textit{computational pangenomics}. 

@@

some modeling and inference topics, related to computational 
systems biology. So, it is my interest to continue my studies with the Ph.D.
to be able to deepen the computer science potential in the biological

@@

Summarizing, the most important research goal of my Ph.D. is designing and

@@

  haplotype sequences as a 
  pangenome graphs, in a compact way via the \textbf{Graph BWT (\textit{GBWT})},
  even if it is more a \textit{multi-string BWT} than a classical graph. A first

@@

  for the \textit{GBWT}, because it would be possible to

@@

  \textit{BIAS} laboratory's topics of interest. So far there has been talk of

@@

  % and develop during my Ph.D.. The management of missing and

@@

some technologies that will be used during my Ph.D. must be mentioned. 

@@

It is therefore my interest to contribute to the development of such open
source libraries. Another point of interest is the

@@

(also on GPU), that are increasingly in use in both \textit{bioinformatics} and
\textit{computational systems biology}.  
\gdv{Secondo me bisogna estendere la parte di ricerca, mettendo qualcosa di più
  a lungo termine. Ad esempio pensare a tecniche di indicizzazione diverse, non
  necessariamente basate su BWT e bitvectors} 

@@

>>>>

\title{\vspace{-1.75cm}PhD Proposal}
\author{Davide Cozzi, 829827,
  \href{mailto:d.cozzi@campus.unimib.it}{d.cozzi@campus.unimib.it}} 
\date{}
\makeatletter
\renewcommand{\paragraph}{%
  \@startsection{paragraph}{4}%
  {\z@}{0.75ex \@plus 1ex \@minus .2ex}{-1em}%
  {\normalfont\normalsize\bfseries}%
}
\makeatother

@@

in a large collection of individual genomes. This new reference, which
was firstly introduced in 2005 for dealing with the comparison of genes in
bacterial species, is called a \emph{pangenome}.  

@@

thousand individuals, in the next few years, also thanks to the latest
developments in sequencing technologies, (\textbf{Next Generation Sequencing

@@

has practical and efficient linear-time solutions if there is proper indexing

@@

  transform (\textit{BWT})}, introduced in 1994 to compress texts, on which the
\textit{FM-index} is based.  

@@

This aligner is based on the \textit{seed-and-extend paradigm}, where pattern

@@

% In recent years, there has been an important development
% in the field of bioinformatics.
As introduced, in the last few years the researchers are focusing on the study
of the \textit{pangenome}, which is a compact 

@@

addition to the specificity of a single genome, allowing even more the use of

@@

an approach that will become increasingly important even in near future.\\ 

@@

The main focus of this proposal is improving the current pattern matching
algorithms and data structures state of the art to represent and query

@@

such data. Theese research fields will have an impact on the ability of

@@

genomic information from \textit{pangenome graphs}. 

@@

% In this context, my PhD is going to be focused on the development of new

@@

A \textit{bitvector}, of the most important \textit{succint data
  structures}, is an array on $n$ bits that allows two particular

@@

1 present in the \textit{bitvector}. Formally, given a bitvector $B$, such

@@

for the analysis of biological sequences.

@@

% in 1994 to compress texts but it has been used widely
% \textit{FM-index}.

@@

type of sequences. A simple but effective strategy to compress texts is based

@@

to introduce some definitions. First of all, the main purpose of these
algorithms is to compute \textbf{Maximal Exact Matches (\textit{MEMs})} of a
pattern in a text. Given a text $T$ and a pattern $P$, a substring of the
pattern $P[i : i+l-1]$, of length $l$, is a \textit{MEM} of $P$ in $T$ if
$P[i:i+l-1]$ is a substring of $T$ but neither $P[i-1:i+l-1]$ nor $P[i:i+l]$
are, that is the match cannot be extended in any direction. The authors state
that computing \textit{MEMs} is equivalent to computing \textbf{matching
  statistics (\textit{MS})}. Formally, the \textit{MS} of a pattern $P$ in
respect to a text $T$ is an array $M$ of pairs position/length ($pos$/$len$),
$|M|=|P|$, such that $T[M[i].pos:M[i].pos+M[i].len-1]=P[i:i+M[i].len]-1]$ and
$P[i:i+M[i].len]$ does not occur in $T$. To compute \textit{MS},
\textit{random-access} over the text is required  but it is also required that
the text is stored in a compressed way. For this purpose, the authors get a
compressed representation of the text, via a \textbf{straight-line program

@@

consecutive runs of the same character. This solution requires two sweeps over
the 

@@

\textit{MS}, through a single sweep and reducing the number of accesses to the
\textit{SLP}. Formally, given two positions $i$ and $j$ in a 

@@

extension between two positions in the text. Summing up, after computing the
\textit{MS array}, as stated by Bannai et al., it is possible to obtain directly

@@

in $SA$, the previous and next index stored in the suffix array. It is important
to note that the use of \textit{bitvectors}, to efficiently store and query 
\textit{runs}, allows many improvements
in memory usage.  In fact, for example, according to Rossi et al., 

@@

biology, have led to the development of algorithms and data
structures for analyzing specific sequence data, such as \textit{haplotypes} and
\textit{genotypes}, in the \textbf{genotyping variants problem}. Briefly, we

@@

the genotype is their combined information. Since 2005, the publication of

@@

example, to study the outcomes of diseases. In this particular historical

@@

columns, not extendable in any direction, such that there is a subset of
identical rows), for the identification of genomic regions that show, for
example, signatures of natural selection.

@@

research based on it, both in terms of variants design, such as the
\textbf{multiallelic PBWT} or the \textbf{dynamic PBWT} (where the static PBWT

@@

structure via linked lists), and in terms of its use to study haplotype panels.
% A first example could be found in the paper of Alanko et al., published in
% 2021, where the authors used the \textit{PBWT} to look for \textit{maximal
%   perfect haplotype block}, within a haplotypes panel. These types of
% algorithms are essential for the identification of genomic regions that show
% signatures of natural selection.
An interesting paper is the one by Williams and Mumey, published in 2020,
who studied \textit{maximal perfect haplotype blocks} with the

@@

software is to make \textit{genotype imputation} to predict unobserved

@@

\textit{GWAS} on cancer and other disorders.\\
In conclusion, as haplotypes panel can be thought of as a particular case of  
\textit{pangenome} (with fixed-length genomes) we can understand the importance
of further improvements to this data structure in \textit{computational
  pangenomics}, especially with regard to space complexity. 

@@

some modelling and inference topics, related to computational 
systems biology. So, it is my interest to continue my studies with PhD
to be able to deepen computer science potential in the biological

@@

In summary, the most important research goal of my PhD is designing and

@@

  haplotype sequences as 
  \textit{pangenome graphs}, in a compact way via the \textbf{Graph BWT
    (\textit{GBWT})}, even if it is more a \textit{multi-string BWT} than a
  classical graph. A first 

@@

  for the \textit{GBWT} because it would be possible to

@@

  \textit{BIAS} laboratory's topics of interest. So far there has been talked of

@@

  % and develop during my PhD. The management of missing and

@@

some technologies that will be used during my PhD must be mentioned. 

@@

It is therefore my interest to contribute to the development of such open-source
libraries. Another point of interest is the 

@@

(also on \textit{GPU}), that are increasingly in use in both
\textit{bioinformatics} and \textit{computational systems biology}.

@@

% \gdv{Secondo me bisogna estendere la parte di ricerca, mettendo qualcosa di più
%   a lungo termine. Ad esempio pensare a tecniche di indicizzazione diverse, non
%   necessariamente basate su BWT e bitvectors} 
